Equalizer Programming Guide

### 2008-10-14 Developer/ProgrammingGuide/paper.tex
Programming Guide: Update for net::Object cleanups
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.10 for Equalizer 0.5.5\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
    1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
    1.3.10  & Oct 7, 2008  & Updated to reflect 2D pixel compounds\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

The destination channel can also be used as a source channel. Contrary
to the other compound modes, it also has to use an output and
corresponding input frame. During rendering, the frustum is 'squeezed'
to configure the pixel decomposition. The destination channel can
therefore not be rendered in place, like with the other compound modes.
\begin{figure}[ht!]\center
  \includegraphics[width=.618\textwidth]{images/pixelKernels.pdf}
  {\caption{\small\label{fPixelKernels}Example pixel kernels for a
      four-to-one pixel compound}}
\end{figure}

@@

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination's channel viewport, frustum and database. A
\textsf{viewport} decomposes the destination channel and frustum in
screen space. A \textsf{range} tells the application to render a part of
its database, and an \textsf{eye} rendering pass can selectively render
different stereo passes. A \textsf{pixel} parameter adapts the frustum
so that the source channel renders every n$th$ line out of m
lines. Setting one or multiple attributes causes the parent's view to be
decomposed accordingly. Attributes are cumulative, that is, intermediate
compound attributes affect and therefore decompose the rendering of all
their children.

@@

Equalizer provides distributed objects which help implementing data
distribution in a cluster environment. The master version of a distributed
object is registered with a \textsf{eq::net::Session}, which assigns a
session-unique identifier to the object. Other nodes can map their
instance of the object to this identifier, thus synchronizing the
object's data with the remotely registered master version.

@@

A static distributed object either has to provide a pointer and size to
its data using \textsf{setInstanceData}, or it has to implement
\textsf{getInstanceData} and \textsf{applyInstanceData}. The first
approach can be used if all distributed member variables are stored in
one contiguous block of memory. The second approach is used otherwise.

The \textsf{InitData} class contains a string of variable
length. Therefore it uses the second approach of manually serializing
and de-serializing its data. Serialization in \textsf{getInstanceData}
and de-serialization in \textsf{applyInstanceData} is performed by
streaming all member variable to or from the provided data
streams. Efficient buffering and data transport between nodes is
implemented in the data streams:

@@

Versioned objects have to override \textsf{getChangeType} to indicate
how they want to have changes to be handled. The current
implementation has the following characteristics:
\begin{itemize}
\item Only the master instance of the object is writable, that is,
  \textsf{eq::net::Object::com\-mit} can be called only on the master
  instance to generate a new version.
\item Slave instance versions can only be advanced, that is,
  \textsf{eq::net::Object::sync( version )} with a version smaller than
  the current version will fail.
\item Newly mapped slave instance are mapped to the oldest available
  version.
\end{itemize}

@@

In addition to the instance data (de)serialization methods used to map
an object, versioned objects may implement \textsf{pack} and
\textsf{unpack} to serialize or de-serialize the changes since the last
version.

If the delta data happens to be layed out contiguously in memory,
\textsf{setDeltaData} might be used. The default implementation of
\textsf{pack} and \textsf{unpack} (de)serialize the delta data or the
instance data if no delta data has been specified.

The \textsf{eqPly} frame data is layed out in one anonymous structure in
memory. It also does not track changes since it is relatively small in
size and changes frequently. Therefore, the instance and delta data are
the same and set in the constructor. Furthermore, the change type is
\textsf{INSTANCE}. The default \textsf{Object} implementation will take
care of the distribution of the data:

{\footnotesize\begin{lstlisting}
        FrameData()
            {
                reset();
                setInstanceData( &data, sizeof( Data ));
                EQINFO << "New FrameData " << std::endl;
            }
\end{lstlisting}}%>>

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-2em}\includegraphics[width=.382\textwidth]{images/modelDist.pdf}
  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
  \vspace{-2em}
\end{wrapfigure}
Some applications might rely on a shared filesystem to access the data,
for example when out-of-core algorithms are used. Other applications
prefer to load the data only on the application process, and use
distributed objects to synchronize the scene data with the render
clients.

@@

The identifer of the root object of this distributed hierarchy is passed
as part of the \textsf{InitData}, and used during remote node
initialization to map the model. \fig{fModelDist} shows the UML
hierarchy of the model and distribution classes.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Scene Graph Distribution}}
\end{wrapfigure}
Each \textsf{VertexBufferDist} object has one node of the model's data
tree. It is serializing the data for this node. Furthermore, it mirrors
the kD-tree by having a \textsf{VertexBufferDist} child for each child
of its corresponding tree node. During serialization, the identifier of
these children is sent to the remote nodes, which reconstruct the mirror
distribution hierarchy and model data tree based on this data.

@@

Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
\end{wrapfigure}
The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} calls first
\textsf{configInitOSWindow}, which creates and initializes the
\textsf{OSWindow} for this window. The \textsf{OSWindow} initialization
code is implementation specific. If the \textsf{OSWindow} was
initialized successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL state
initialization. The default implementation sets up some typical OpenGL
state, e.g., it enables the depth test.

The \textsf{OSWindow} initialization takes into account various
attributes set in the configuration file. Attributes include the size of
the various frame buffer planes (color, alpha, depth, stencil) as well
as other framebuffer attributes, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case the Equalizer default implementation
will test for their availability and enable them if possible.

@@

The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
\end{wrapfigure}
The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visiblity against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
The \textsf{OSWindow} implementation is responsible for setting up and
de-initializing event handling.

Carbon events issued for AGL windows are processed from the process main
thread. One \textsf{AGLEventHandler} is used. When a Carbon window is
set on an \textsf{AGLWindow}, the window is registered with the
\textsf{AGLEventHandler}. The event handler installs a Carbon event
handler for all important events. The event handler uses an
\textsf{AGLWindowEvent} to pass the Carbon \textsf{EventRef} to
\textsf{AGLWindowIF::processEvent}. During window exit, the installed
Carbon handler is removed when the window is deregistered with the
\textsf{AGLEventHandler}. No event handling is set up for PBuffers.

@@

The glX event handler iterates over all windows of the registered pipe
to find the originating \textsf{GLXWindowIF}, which is then used to
process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::processEvent}.

Each \textsf{WGLWindow} allocates one \textsf{WGLEventHandler} when the
window handle is set. The WGLEventHandler passes the native event
parameters \textsf{uMsg}, \textsf{wParam} and \textsf{lParam} to
\textsf{WGLWindowIF::processEvent} as part of the
\textsf{WGLWindowEvent}. No event handling is set up for PBuffers.

@@

The receiver and command thread are terminated when the application
stops network communications using \textsf{Client::exitLocal}.

@@

The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entities
unlock all their children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

This section explains window initialization in detail. It discusses in
detail the handling on the different window systems. The entry point for
the window initialization is the task method
\textsf{Window::configInit}. This task method first calls a
window-system-specific task method, and then
\textsf{Window::configInitGL} to do generic OpenGL state setup.

Since window initialization is notoriously error-prone and hard to
debug, the default Equalizer functions propagate the reason for errors
from the render clients back to the application. The \textsf{Pipe} and
\textsf{Window} classes have a \textsf{setErrorMessage} method, which is
used to set an error string. This string is passed to the
\textsf{Config} instance on the application node, where it can be
queried using \textsf{getErrorMessage}.

The window-system-specific initialization methods use overrideable
methods for all sub-tasks. This allows partial customization, without
the need of rewriting tedious window initialization code, e.g., the
OpenGL pixel format selection.

@@

The pixel format is chosen based on the window's attributes. Some
attributes set to auto, e.g., stereo, cause the method first to request
the feature and then to back off and retry if it is not available. The
returned pixel format has to be destroyed using
\textsf{Window::destroyAGLPixelFormat}. When no matching pixel format is
found, \textsf{chooseAGLPixelFormat} returns \textsf{0} and the AGL
window initialization returns with a failure.

@@

{\footnotesize\begin{lstlisting}
bool Window::configInitAGL()
{
    AGLPixelFormat pixelFormat = chooseAGLPixelFormat();
    if( !pixelFormat )
        return false;

@@

{\footnotesize\begin{lstlisting}
bool Window::configInitGLX()
{
    XVisualInfo* visualInfo = chooseXVisualInfo();
    if( !visualInfo )
        return false;

@@

    if( success && !_xDrawable )
    {
        setErrorMessage( "configInitGLXDrawable did set no X11 drawable" );
        return false;
    }

@@

\begin{enumerate}
\item\textsf{getWGLPipeDC} is used to get an affinity device context,
  which might be needed for window creation. If a context is returned, a
  function pointer to delete the returned device context is set as
  well. This function might return 0, which is not an error.
\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
  window attributes. If no device context is given, it uses the system
  device context. The chosen pixel format is set on the passed device
  context.
\item\textsf{configInitWGLDrawable} creates the drawable. The device
  context passed to \textsf{configInitWGLDrawable} is used to query the
  pixel format and is used as the device context for creating a
  PBuffer. If no device context is given, the display device context is
  used. On success, it sets the window handle. Setting a window handle
  also sets the window's device context.
\item\textsf{createWGLContext} creates an OpenGL rendering context using
  the given device context. If no device context is given, the window's
  device context is used. This function does not set the window's OpenGL
  context.
\end{enumerate}

@@

{\footnotesize\begin{lstlisting}
bool Window::configInitWGL()
{
    PFNEQDELETEDCPROC deleteDCProc = 0;
    HDC dc = getWGLPipeDC( deleteDCProc );
    EQASSERT( !dc || deleteDCProc );

@@

    if( !_wglDC )
    {
        if( dc )
            deleteDCProc( dc );
        setErrorMessage( "configInitWGLDrawable did not set a WGL drawable" );
        return false;
    }

@@

@@

    if( dc )
        deleteDCProc( dc );
    return true;
}
\end{lstlisting}}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
      Volume Rendering}}
\end{wrapfigure}
 When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependent data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
  \vspace{-1em}
\end{wrapfigure}
For recomposition, the 2D frame buffer contents are blended together to
form a seamless picture. For correct blending, the frames are ordered in
the same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.11 for Equalizer 0.5.5+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
    1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
    1.3.10  & Oct 7, 2008  & Updated to reflect 2D pixel compounds\\
    1.3.11  & Oct 13, 2008 & Updated to reflect eq::net::Object cleanup\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

The destination channel can also be used as a source channel. Contrary
to the other compound modes, it also has to use an output and
corresponding input frame. During rendering, the frustum is 'squeezed'
to configure the pixel decomposition. The destination channel can
therefore not be rendered in place, like with the other compound modes.
\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/pixelKernels.pdf}
  {\caption{\small\label{fPixelKernels}Example pixel kernels for a
      four-to-one pixel compound}}
\end{figure}

@@

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination's channel viewport, frustum and database. A
\textsf{viewport} decomposes the destination channel and frustum in
screen space. A \textsf{range} tells the application to render a part of
its database, and an \textsf{eye} rendering pass can selectively render
different stereo passes. A \textsf{pixel} parameter adapts the frustum
so that the source channel renders an evenly subset of the parent's
pixels. Setting one or multiple attributes causes the parent's view to
be decomposed accordingly. Attributes are cumulative, that is,
intermediate compound attributes affect and therefore decompose the
rendering of all their children.

@@

Equalizer provides distributed objects which help implementing data
distribution in a cluster environment. The master version of a
distributed object is registered with a \textsf{eq::net::Session}, which
assigns a session-unique identifier to the object. This identifier can
be used by other nodes to map their instance of the object, thus
synchronizing the object's data with the remotely registered master
version.

@@

A static distributed object has to implement \textsf{getInstanceData}
and \textsf{applyInstanceData} to serialize and deserialize the object's
distributed data. These method provide an output or input stream as a
parameter, which abstract the data transmission and can be used like a
\textsf{std::stream}.

For \textsf{InitData}, serialization in \textsf{getInstanceData}
and de-serialization in \textsf{applyInstanceData} is performed by
streaming all member variables to or from the provided data
streams. Efficient buffering and data transport between nodes is
implemented in the data streams:

@@

Versioned objects have to override \textsf{getChangeType} to indicate
how they want to have changes to be handled. All versioned objects
currently implemented have the following characteristics:
\begin{itemize}
\item Only the master instance of the object is writable, that is,
  \textsf{eq::net::Object::com\-mit} can be called only on the master
  instance to generate a new version.
\item Slave instance versions can only be advanced, that is,
  \textsf{eq::net::Object::sync( version )} with a version smaller than
  the current version will fail.
\item Newly mapped slave instance are mapped to the oldest available
  version by default, or to the specified version in
  \textsf{Session::mapObject}.
\end{itemize}

@@

In addition to the instance data (de-)serialization methods used to map
an object, versioned objects may implement \textsf{pack} and
\textsf{unpack} to serialize or de-serialize the changes since the last
version. If these methods are not implemented, their default
implementation forwards the (de-)serialization request to
\textsf{getInstanceData} and \textsf{applyInstanceData}, respectively.

The \textsf{eqPly} does not track changes since it is relatively small
in size and changes frequently. Therefore, the instance and delta data
are the same, and \textsf{pack} and \textsf{unpack} are not
implemented. The change type is \textsf{INSTANCE}. The serialization
functions exploit the fact that the frame data is layed
out in one anonymous structure in memory, and transmit all the data in
one call:

{\footnotesize\begin{lstlisting}
void FrameData::getInstanceData( eq::net::DataOStream& os )
{ 
    os.writeOnce( &data, sizeof( data )); 
}

void FrameData::applyInstanceData( eq::net::DataIStream& is )
{
    memcpy( &data, is.getRemainingBuffer(), sizeof( data ));
    is.advanceBuffer( sizeof( data ));
}
\end{lstlisting}}%>>

@@

Some applications might rely on a shared filesystem to access the data,
for example when out-of-core algorithms are used. Other applications
prefer to load the data only on the application process, and use
distributed objects to synchronize the scene data with the render
clients.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
\end{wrapfigure}
The identifer of the root object of this distributed hierarchy is passed
as part of the \textsf{InitData}, and used during remote node
initialization to map the model. \fig{fModelDist} shows the UML
hierarchy of the model and distribution classes.

Each \textsf{VertexBufferDist} object has one node of the model's data
tree. It is serializing the data for this node. Furthermore, it mirrors
the kD-tree by having a \textsf{VertexBufferDist} child for each child
of its corresponding tree node. During serialization, the identifier of
these children is sent to the remote nodes, which reconstruct the mirror
distribution hierarchy and model data tree based on this data.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Scene Graph Distribution}}
\end{wrapfigure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} calls first
\textsf{configInitOSWindow}, which creates and initializes the
\textsf{OSWindow} for this window. The \textsf{OSWindow} initialization
code is implementation specific. If the \textsf{OSWindow} was
initialized successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL state
initialization. The default implementation sets up some typical OpenGL
state, e.g., it enables the depth test.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
\end{wrapfigure}
The \textsf{OSWindow} initialization takes into account various
attributes set in the configuration file. Attributes include the size of
the various frame buffer planes (color, alpha, depth, stencil) as well
as other framebuffer attributes, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case the Equalizer default implementation
will test for their availability and enable them if possible.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
  \vspace{-5ex}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree During
this preprocessing step, each node of the tree gets a database range
assigned. The root node has the range [0, 1], its left child [0, 0.5]
and its right child [0.5, 1], and so on for all nodes in the tree. The
preprocessed model is saved in a binary format for accelerating
subsequent loading.

The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visiblity against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

The concrete implementation of these window interfaces is responsible
for setting up and de-initializing event handling.

Carbon events issued for AGL windows are processed from the node's main
thread. One \textsf{AGLEventHandler} per process is used. When a Carbon
window is set on an \textsf{AGLWindow}, the window is registered with
the global \textsf{AGLEventHandler}. The event handler installs a Carbon
event handler for all important events. The event handler uses an
\textsf{AGLWindowEvent} to pass the Carbon \textsf{EventRef} to
\textsf{AGLWindowIF::processEvent}. During window exit, the installed
Carbon handler is removed when the window is deregistered with the
\textsf{AGLEventHandler}. No event handling is set up for PBuffers.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
The glX event handler iterates over all windows of the registered pipe
to find the originating \textsf{GLXWindowIF}, which is then used to
process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::\-process\-Event}.

Each \textsf{WGLWindow} allocates one \textsf{WGLEventHandler} when the
window handle is set. The WGLEventHandler passes the native event
parameters \textsf{uMsg}, \textsf{wParam} and \textsf{lParam} to
\textsf{WGL\-Win\-dowIF::processEvent} as part of the
\textsf{WGLWindowEvent}. No event handling is set up for PBuffers.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\small\label{fThreads}Threads within one node process}}
\end{wrapfigure}
The receiver and command thread are terminated when the application
stops network communications using \textsf{Client::exitLocal}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entities
unlock all their children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

This section explains window initialization in detail. It discusses in
detail the handling of the different window systems. Window systems are
integrated through the \textsf{OSWindow} interface, which externalizes
the details of the windowing API from the \textsf{eq::Window}
implementation and facilitates the integration with new windowing API's
and custom applications.

Equalizer provides sample implementations for the supported window
systems. These sample implementations are intended to be sub-classed and
various steps in the window initialization can be overwritten and
customized.

An application typically choses to subclass the sample implementation if
only minor tweaks are needed for integration. For major changes or new
window systems, it is often easier to subclass directly from
\textsf{OSWindow} and implement the abstract methods of this interface
class.

The method \textsf{Window::configInitOSWindow} is used to instantiate
and initialize the \textsf{OSWindow} implementation during config
initialization. After an successful \textsf{OSWindow} initialization,
\textsf{Window::configInitGL} is called for the generic OpenGL state
setup.

Since window initialization is notoriously error-prone and hard to
debug, the sample implementation propagates the reason for errors
from the render clients back to the application. The \textsf{Pipe} and
\textsf{Window} classes have a \textsf{setErrorMessage} method, which is
used to set an error string. This string is passed to the
\textsf{Config} instance on the application node, where it can be
queried using \textsf{getErrorMessage}.

The sample implementations \textsf{AGLWindow}, \textsf{GLXWindow} and
\textsf{WGLWindow} all have similar, overrideable methods for all
sub-tasks. This allows partial customization, without the need of
rewriting tedious window initialization code, e.g., the OpenGL pixel
format selection.

@@

The pixel format is chosen based on the window's attributes. Some
attributes set to auto, e.g., stereo, cause the method first to request
the feature and then to back off and retry if it is not available. The
pixel format returned by \textsf{chooseAGLPixelFormat} has to be
destroyed using \textsf{destroyAGLPixelFormat}. When no matching pixel
format is found, \textsf{chooseAGLPixelFormat} returns \textsf{0} and
the AGL window initialization returns with a failure.

@@

{\footnotesize\begin{lstlisting}
bool AGLWindow::configInit( )
{
    AGLPixelFormat pixelFormat = chooseAGLPixelFormat();
    if( !pixelFormat )
        return false;

@@

{\footnotesize\begin{lstlisting}
bool GLXWindow::configInit( )
{
    XVisualInfo* visualInfo = chooseXVisualInfo();
    if( !visualInfo )
        return false;

@@

    if( success && !_xDrawable )
    {
        _window->setErrorMessage( 
            "configInitGLXDrawable did not set a X11 drawable");
        return false;
    }

@@

\begin{enumerate}
\item\textsf{getWGLPipeDC} is used to get an affinity device context,
  which might be needed for window creation. If a context is returned, a
  function pointer to delete the returned device context is set as
  well. This function might return 0, which is not an error, but
  signifies that no affinity context is used.
\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
  window attributes. If no device context is given, it uses the system
  device context. The chosen pixel format is set on the passed device
  context.
\item\textsf{configInitWGLDrawable} creates the drawable. The device
  context passed to \textsf{configInitWGLDrawable} is used to query the
  pixel format and is used as the device context for creating a
  PBuffer. If no device context is given, the display device context is
  used. On success, it sets the window handle. Setting a window handle
  also sets the window's device context.
\item\textsf{createWGLContext} creates an OpenGL rendering context using
  the given device context. If no device context is given, the window's
  device context is used. This function does not set the window's OpenGL
  context.
\end{enumerate}

@@

{\footnotesize\begin{lstlisting}
bool WGLWindow::configInit()
{
    PFNEQDELETEDCPROC deleteDCProc = 0;
    HDC dc = getWGLPipeDC( deleteDCProc );
    EQASSERT( !dc || deleteDCProc );

@@

    if( !_wglDC )
    {
        if( dc )
            deleteDCProc( dc );
        _window->setErrorMessage(
            "configInitWGLDrawable did not set a WGL drawable" );
        return false;
    }

@@

    if( getIAttribute( Window::IATTR_HINT_SWAPSYNC ) != AUTO )
    {
        if( WGLEW_EXT_swap_control )
        {
            // set vsync on/off
            const GLint vsync = 
                ( getIAttribute( Window::IATTR_HINT_SWAPSYNC )==OFF ) ? 0 : 1;
            wglSwapIntervalEXT( vsync );
        }
        else
            EQWARN << "WGLEW_EXT_swap_control not supported, ignoring window "
                   << "swapsync hint" << std::endl;
    }

@@

    if( dc )
        deleteDCProc( dc );
    return true;
}
\end{lstlisting}}%>>

@@

When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
      Volume Rendering}}
\end{wrapfigure}
DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependent data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

For recomposition, the 2D frame buffer contents are blended together to
form a seamless picture. For correct blending, the frames are ordered in
the same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

\begin{figure}[h!t]
  \subfigure[]{\label{fBlendOrtho}
    \includegraphics[width=.45\textwidth]{images/b2f_ortho.pdf}
  }\hfil
  \subfigure[]{\label{fBlendPerspective}
    \includegraphics[width=.55\textwidth]{images/b2f_perspective.pdf}
  }%
  {\caption{\small\label{fBlend}Back-to-Front Com\-po\-siting for
      Orthogonal and Perspective View Frustra}}
\end{figure}

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
  \vspace{-2ex}
\end{wrapfigure}
For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.
